---
title: "Accumulated Cyclone Energy (ACE)"
author: "Tim Trice"
date: "June 27, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      fig.align = "center")
```

```{r, message = FALSE}
library(dplyr)
library(ggplot2)

# library(HURDAT)
library(lubridate)
library(readr)
library(rrricanes)
#library(rrricanesdata)
```

ACE or Accumulated Cyclone Energy is a method of measuring energy of a cyclone or for an entire season. It is calculated by the formula

$ \text{ACE} = 10^{-4}\sum{v^2_\text{max}} $

where $v_\text{max}$ is the wind speed in knots. Values may only be used when a storm is a tropical system with winds of at least 35 knots. Additionally, only six-hour intervals are used.

To calculate ACE you would want to use the `fstadv` dataset and apply the following rules:

* since forecast/advisory products are typically issued at 03:00, 09:00, 15:00 and 21:00 UTC filter out odd hours
* `Status` is Tropical Storm or Hurricane.
* `Wind` is not NA
* group by `StormKey`
* select `Wind`

```{r eval=FALSE}

fstadv <- fstadv |> 
    filter(hour(Date) %in% c(3, 9, 15, 21), 
           Status %in% c("Tropical Storm", "Hurricane"), 

           !is.na(Wind)) |>
    group_by(StormKey) |> 

    select(Name, Wind)
```

Now let's summarise our dataset with new variable `ACE`.

```{r eval=FALSE}
fstadv |> 
    summarise(Name = last(Name), 
              ACE = sum(Wind^2) * 1e-04) |> 
    arrange(desc(ACE)) |> 
    top_n(10)
```

This matches somewhat well with [Wikipedia](https://en.wikipedia.org/wiki/Accumulated_cyclone_energy#Individual_storms) and other sources. But, you may notice we're missing some storms. `rrricanes` currently only holds data back to 1998; this data is considered "real-time". 

A companion package, [HURDAT](https://github.com/timtrice/HURDAT) is available in CRAN that has data for all cyclones dating back as far as 1851. This package has less data than `rrricanes`. But, as it is based on a post-storm reanalysis project, the data is more accurate. 

Let's revisit the top 10 using `HURDAT`:
Note:HURDAT is currently archived on CRAN.


```{r eval=FALSE}

AL |> 
    filter(hour(DateTime) %in% c(0, 6, 12, 18), 
           Status %in% c("TS", "HU"), 

           !is.na(Wind)) |> 
    group_by(StormKey) |>

    summarise(Name = last(Name), 
              ACE = sum(Wind^2) * 1e-04) |> 
    arrange(desc(ACE)) |> 
    top_n(10)
```

A couple of things to notice here:

1. in `HURDAT`, the common times used are 00:00, 06:00, 12:00 and 18:00 UTC
2. Our list is more comprehensive than the Wikipedia list as that list only measures storms after 1950. 

`ACE` is slightly higher and that could be for a number of reasons. For example, on re-analysis the Hurricane Research Division may have determined a cyclone was actually tropical (shown in `HURDAT`) when initially it was believed to be extratropical (as shown in `rrricanes`). Or, and more likely, they determined through additional data that a storm was actually stronger than originally though. 

You can also calculate `ACE` for a season. Instead of grouping by `StormKey` we group by `Year`. I'll stick with `HURDAT` in this example.

```{r eval=FALSE}
(df <- AL |> 
    mutate(Year = year(DateTime)) |> 
    filter(hour(DateTime) %in% c(0, 6, 12, 18), 
           Status %in% c("TS", "HU"), 
           !is.na(Wind)) |> 
    group_by(Year) |> 
    summarise(ACE = sum(Wind^2) * 1e-04) |> 
    arrange(desc(ACE))) |> 
    top_n(10)
```

This also matches relatively well with that on [Wikipedia](https://en.wikipedia.org/wiki/Accumulated_cyclone_energy#Atlantic_hurricane_seasons.2C_1950.E2.80.932017) and other sources. 

```{r eval=FALSE}
ggplot(df, aes(x = Year, y = ACE)) + 
    geom_bar(stat = "identity") + 
    theme_bw()
```

It would certainly seem that tropical cyclone activity ebbs and flows over time. 
