---
title: "GIS data"
author: "Tim Trice"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GIS data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, message = FALSE}
library(dplyr)
library(ggplot2)
library(rrricanes)
```

Most storms will contain a variation of GIS datasets that can be plotted with `ggplot2`. The helper functions for this have the prefix 'gis'.

**All products are experimental and there maybe fluctuations particularly in current datasets.**

In general, datasets are available for storms dated back to 1998. However, products such as Wind Speed Probabilities only go back to 1999. 

Some datasets require the use of the storm key and an optional advisory. Other products require a datetime value and cannot be isolated by storm key or advisory. The datetime values are not based on the issue time of the advisory, but rather three hours prior. For example, if you are seeking a dataset where the forecast/advisory was issued at 9:00AM UTC, you will want the dataset for 6:00AM UTC. This will be explained a little further below.

## Build a Tracking Chart

There are three functions available to help you plot GIS data; `tracking_chart`, `al_tracking_chart` and `ep_tracking_chart`. `al_tracking_chart` and `ep_tracking_chart` are just helpers centered on the Atlantic and northeast Pacific ocean, respectively. 

Building tracking charts requires the `rnaturalearthdata` and, for high resolution charts, `rnaturalearthhires` packages.

```{r}
args(rrricanes::tracking_chart)
```

The countries and states parameters are TRUE by default. This means a basic call to `tracking_chart` will return a map with country and state borders. The basic call does not show this.

```{r, fig.width = 7, fig.align = "center"}
tracking_chart()
```

You can pass typical `aes` parameters to refine the color and fill of the plot; remember the tracking chart is a ggplot object.

```{r, fig.width = 7, fig.align = "center"}
tracking_chart(color = "black", fill = "white", size = 0.1)
```

You may choose to only show coastline data instead. In this case, just set the countries parameter to FALSE.

```{r, fig.width = 7, fig.align = "center"}
tracking_chart(countries = FALSE, res = 50, color = "black", fill = "white", 
               size = 0.1)
```

For the purposes of this vignette we'll focus on Atlantic storms.

```{r, fig.width = 7, fig.align = "center"}
(p <- al_tracking_chart(color = "black", fill = "white", size = 0.1, res = 50))
```

The `res` parameter defines the resolution of the chart presented. Options are in 110nm, 50nm and 10nm. The lower the resolution the longer the chart takes to be built. 

States cannot be drawn on resolution greater than 50nm. 

## GIS Datasets

There are several datasets that are published for active cyclones. The following functions are designed to return the URL to those datasets:

* `gis_advisory`
* `gis_prob_storm_surge`
* `gis_windfield`
* `gis_latest`

## Advisory Package

```{r}
gis_advisory(key = "AL182012", advisory = "18")
```

```{r, eval = FALSE}
adv <- gis_advisory(key = "AL182012", advisory = "18") %>% gis_download()
```

```{r}
names(adv)
```

For this particular storm and advisory, included are the base line, point and polygon datasets along with a dataset for watches and warnings. The objects returned are spatial dataframes contained within the list of dataframes, `adv`.

### Line track

```{r}
str(adv$al182012_018_5day_lin)
```

```{r}
fcst_line <- as(adv$al182012_018_5day_lin, "SpatialPointsDataFrame") %>%
    broom::tidy()
```

```{r, fig.width = 7, fig.align = "center"}
p + geom_path(data = fcst_line, aes(coords.x1, coords.x2, group = FCSTPRD))
```

### Point track

```{r, fig.width = 7, fig.align = "center"}
p + geom_point(data = adv$al182012_018_5day_pts@data, aes(LON, LAT))
```

## Forecast Cone

Forecast cone data is contained in the polygon dataset. To deal with this dataset we need to make some steps to build an appropriate dataframe.

```{r}
fcst <- adv$al182012_018_5day_pgn
fcst@data$id <- rownames(fcst@data)
fcst.points = broom::tidy(fcst, region = "id")
fcst <- dplyr::left_join(fcst.points, fcst@data, by = "id")
```

```{r, fig.width = 7, fig.align = "center"}
p + geom_polygon(data = fcst %>% filter(FCSTPRD == 120), 
                 aes(long, lat, group = group, fill = factor(FCSTPRD))) + 
    geom_polygon(data = fcst %>% filter(FCSTPRD == 72), 
                 aes(long, lat, group = group, fill = factor(FCSTPRD)))
```

## Watches and Warnings

```{r, fig.width = 7, fig.align = "center"}
p + geom_path(data = adv$al182012_018_ww_wwlin, 
               aes(x = long, y = lat, group = group), size = 2)
```

## Probabilistic Storm Surge

The Tropical Cyclone Storm Surge Probabilities data shows the probability, in percent, of a specified storm surge occurring during the forecast period indicated. The product is based upon an ensemble of Sea, Lake, and Overland Surge from Hurricanes (SLOSH) model runs using the National Hurricane Center (NHC) official advisory and accounts for track, size, and intensity errors based on historical errors.

see: http://www.nhc.noaa.gov/gis/metadata_ex_gt5_at.xml

* Does not take an advisory number
* `datetime` and `nobs` parameters created to help

cumulative probablity of storm surge greater than 5ft

```{r}
gis_prob_storm_surge(key = "AL142016", products = list(psurge = 0), 
                     datetime = "20161006")
```

```{r, eval = FALSE}
ss <- gis_prob_storm_surge(key = "AL142016", 
                           products = list(psurge = 0), 
                           datetime = "20161006") %>% 
    last() %>% 
    gis_download()
```

```{r, fig.width = 7, fig.align = "center"}
p + geom_path(data = ss$al142016_2016100618_gt0, 
               aes(x = long, y = lat, group = group), size = 1)
```

## Current and Forecast Wind Field Radii

```{r}
gis_windfield("AL142016", advisory = "33")
```

```{r, eval = FALSE}
wf <- gis_windfield("AL142016", advisory = "33") %>% gis_download()
```

Initial radius

```{r}
wf_init <- wf$al142016_2016100606_initialradii
wf_init@data$id <- rownames(wf_init@data)
wf_init.points = broom::tidy(wf_init, region = "id")
wf_init <- dplyr::left_join(wf_init.points, wf_init@data, by = "id")
```

```{r, fig.width = 7, fig.align = "center"}
p + geom_polygon(data = wf_init, 
                  aes(x = long, y = lat, fill = factor(RADII)), alpha = 0.5)
```

```{r}
wf_fcst <- wf$al142016_2016100606_forecastradii
wf_fcst@data$id <- rownames(wf_fcst@data)
wf_fcst.points = broom::tidy(wf_fcst, region = "id")
wf_fcst <- dplyr::left_join(wf_fcst.points, wf_fcst@data, by = "id")
```

```{r, fig.width = 7, fig.align = "center"}
p + geom_polygon(data = wf_fcst, 
                  aes(x = long, y = lat, group = group, fill = factor(RADII)), 
                  alpha = 0.5)
```

## Wind Speed Probabilities

```{r}
gis_wsp(datetime = "2016100606", res = 0.5)
```

```{r, eval = FALSE}
wsp <- gis_wsp(datetime = "2016100606", res = 0.5) %>% gis_download()
```

### Cumulative Probability for >34kt Winds

```{r, fig.width = 7, fig.align = "center"}
p + geom_point(data = broom::tidy(wsp$`2016100606_wsp34knt120hr_halfDeg`), 
               aes(x = coords.x1, y = coords.x2, color = PWIND120))
```

### Cumulative Probability for >50kt Winds

```{r, fig.width = 7, fig.align = "center"}
p + geom_point(data = broom::tidy(wsp$`2016100606_wsp50knt120hr_halfDeg`), 
               aes(x = coords.x1, y = coords.x2, color = PWIND120))
```

### Cumulative Probability for >64kt Winds

```{r, fig.width = 7, fig.align = "center"}
p + geom_point(data = broom::tidy(wsp$`2016100606_wsp64knt120hr_halfDeg`), 
               aes(x = coords.x1, y = coords.x2, color = PWIND120))
```
