---
title: "Installing rrricanesdata"
author: "Tim Trice"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Installing rrricanesrdata}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r echo = FALSE, message = FALSE, warning = FALSE}
library(rrricanesdata)
library(rrricanes)
```

With `rrricanes` you can access current and archived advisories as-issued from the National Hurricane Center archives. `rrricanesdata` is a complimentary data package meant to make it faster to get most data avialable from these storms. 

Installing `rrricanesdata` is simple:

```{r, eval = FALSE}
install.packages("rrricanesdata", 
                 repos = "https://timtrice.github.io/drat/", 
                 type = "source")
```

Data in `rrricanesdata` will be updated on the first of every month with a cutoff date of midnight on the last day of the month. So, advisories issued at any time during the current month will not be available; you will need to use any of `rrricanes` `get_*` functions. 

## Datasets

### `adv`
  * StormKey: Unique identifier of cyclone
  * Adv: Advisory number
  * Date: Date and time of advisory
  * Status: Classification of cyclone
  * Name: Name of cyclone
  * Lat: Latitude of cyclone center
  * Lon: Longitude of cyclone center
  * Wind: Maximum sustained one-minute winds in knots
  * Gust: Maximum sustained one-minute gusts in knots
  * Pressure: Minimum central pressure in millibars
  * PosAcc: Position accuracy of cyclone in nautical miles
  * FwdDir: Compass angle of forward motion
  * FwdSpeed: Forward speed in miles per hour
  * Eye: Size of eye in nautical miles
  * SeasNE: Radius of 12ft seas in northeast quadrant
  * SeasSE: Radius of 12ft seas in southeast quadrant
  * SeasSW: Radius of 12ft seas in southwest quadrant
  * SeasNW: Radius of 12ft seas in northwest quadrant

### `discus`
  * Status: Classification of storm, e.g., Tropical Storm, Hurricane,
    etc.
  * Name: Name of storm
  * Adv: Advisory Number
  * Date: Date of advisory issuance
  * StormKey: ID of cyclone
  * Contents: Text content of product

### `fcst`
  * StormKey: Unique identifier of cyclone
  * Adv: Advisory number
  * Date: Date and time of advisory
  * FcstDate: Forecast date and time in UTC
  * Lat: Forecast latitude
  * Lon: Forecast Longitude
  * Wind: Forecast wind in knots
  * Gust: Forecast gust in knots

### `fcst_wr`
  * StormKey: Unique identifier of cyclone
  * Adv: Advisory number
  * Date: Date and time of advisory
  * FcstDate: Forecast date and time in UTC
  * WindField: Minimum sustained wind field for quadrants
  * NE: Radius in nautical miles for northeast quadrant
  * SE: Radius in nautical miles for southeast quadrant
  * SW: Radius in nautical miles for southwest quadrant
  * NW: Radius in nautical miles for northwest quadrant

### `fstadv`
  * Status: Classification of cyclone
  * Name: Name of cyclone
  * Adv: Advisory number
  * Date: Date and time of advisory
  * StormKey: Unique identifier of cyclone
  * Lat: Latitude of cyclone center
  * Lon: Longitude of cyclone center
  * Wind: Maximum sustained one-minute winds in knots
  * Gust: Maximum sustained one-minute gusts in knots
  * Pressure: Minimum central pressure in millibars
  * PosAcc: Position accuracy of cyclone in nautical miles
  * FwdDir: Compass angle of forward motion
  * FwdSpeed: Forward speed in miles per hour
  * Eye: Size of eye in nautical miles
  * NE64: Radius of >=64kt winds in northeast quadrant
  * SE64: Radius of >=64kt winds in southeast quadrant
  * SW64: Radius of >=64kt winds in southwest quadrant
  * NW64: Radius of >=64kt winds in northwest quadrant
  * NE50: Radius of >=50kt winds in northeast quadrant
  * SE50: Radius of >=50kt winds in southeast quadrant
  * SW50: Radius of >=50kt winds in southwest quadrant
  * NW50: Radius of >=50kt winds in northwest quadrant
  * NE34: Radius of >=34kt winds in northwest quadrant
  * SE34: Radius of >=34kt winds in southeast quadrant
  * SW34: Radius of >=34kt winds in southwest quadrant
  * NW34: Radius of >=34kt winds in northwest quadrant
  * Hr\{n\}FcstDate: Forecast valid date
  * Hr\{n\}Lat: Forecast latitude in `n` hours
  * Hr\{n\}Lon: Forecast longitude in `n` hours
  * Hr\{n\}Wind: Forecast maximum wind in `n` hours
  * Hr\{n\}Gust: Forecast maximum gust in `n` hours
  * Hr\{n\}NE64: Forecast wind radius in `n` hours
  * Hr\{n\}SE64: Forecast wind radius in `n` hours
  * Hr\{n\}SW64: Forecast wind radius in `n` hours
  * Hr\{n\}NW64: Forecast wind radius in `n` hours
  * Hr\{n\}NE50: Forecast wind radius in `n` hours
  * Hr\{n\}SE50: Forecast wind radius in `n` hours
  * Hr\{n\}SW50: Forecast wind radius in `n` hours
  * Hr\{n\}NW50: Forecast wind radius in `n` hours
  * Hr\{n\}NE34: Forecast wind radius in `n` hours
  * Hr\{n\}SE34: Forecast wind radius in `n` hours
  * Hr\{n\}SW34: Forecast wind radius in `n` hours
  * Hr\{n\}NW34: Forecast wind radius in `n` hours
  * SeasNE: Radius of 12ft seas in northeast quadrant
  * SeasSE: Radius of 12ft seas in southeast quadrant
  * SeasSW: Radius of 12ft seas in southwest quadrant
  * SeasNW: Radius of 12ft seas in northwest quadrant

### `posest`
  * Status: Classification of storm, e.g., Tropical Storm, Hurricane,
    etc.
  * Name: Name of storm
  * Date: Date of advisory issuance
  * Contents: Text content of product

### `prblty`
  * Status: Classification of storm, e.g., Tropical Storm, Hurricane,
    etc.
  * Name: Name of storm
  * Adv: Advisory Number
  * Date: Date of advisory issuance
  * Location: Location for which the probability statistics rely
  * A: Probability of a strike within the next 12 hours
  * B: Probability of a strike between 12 and 24 hours
  * C: Probability of a strike between 24 and 36 hours
  * D: Probability of a strike between 36 and 48 hours
  * E: Probability of a strike between 48 and 72 hours

### `public`
  * Status: Classification of storm, e.g., Tropical Storm, Hurricane,
    etc.
  * Name: Name of storm
  * Adv: Advisory Number
  * Date: Date of advisory issuance
  * StormKey: Unique ID of the cyclone
  * Contents: Text content of product

### `storms`
  * StormKey: Storm ID
  * Name: Storm name
  * Wind: Peak wind speed in knots
  * StartDate: Date/time of first advisory
  * EndDate: Date/time of last advisory

### `update`
  * Status: Classification of storm, e.g., Tropical Storm, Hurricane,
    etc.
  * Name: Name of storm
  * Date: Date of advisory issuance
  * StormKey: Unique ID of cyclone
  * Contents: Text content of product

### `wndprb`
  * Status: Classification of storm, e.g., Tropical Storm, Hurricane,
    etc.
  * Name: Name of storm
  * Adv: Advisory Number
  * Date: Date of advisory issuance
  * Wind: Minimum wind speed for which probabilities reference
  * Wind12: Probability of sustained `Wind` within 12 hours
  * Wind24: Probability of sustained `Wind` within 24 hours
  * Wind24Cum: Cumulative probability through 24 hours
  * Wind36: Probability of sustained `Wind` within 36 hours
  * Wind36Cum: Cumulative probability through 36 hours
  * Wind48: Probability of sustained `Wind` within 48 hours
  * Wind48Cum: Cumulative probability through 48 hours
  * Wind72: Probability of sustained `Wind` within 72 hours
  * Wind72Cum: Cumulative probability through 72 hours
  * Wind96: Probability of sustained `Wind` within 96 hours
  * Wind96Cum: Cumulative probability through 96 hours
  * Wind120: Probability of sustained `Wind` within 120 hours
  * Wind120Cum: Cumulative probability through 120 hours

### `wr`
  * StormKey: Unique identifier of cyclone
  * Adv: Advisory number
  * Date: Date and time of advisory
  * Windfield: Minimum wind speed expected
  * NE: Radius of `Windfield` in the northeast quadrant
  * SE: Radius of `Windfield` in the southeast quadrant
  * SW: Radius of `Windfield` in the southwest quadrant
  * NW: Radius of `Windfield` in the northwest quadrant
