---
title: "Probabilistic Storm Surge"
author: "Tim Trice"
date: "June 11, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      fig.width = 10, 
                      fig.asp = 1, 
                      fig.align = "center")
```

```{r, message = FALSE}
library(dplyr)
library(ggplot2)
library(rrricanes)
library(rrricanesdata)
library(sp)
```

## Data

```{r}
key <- "AL092008"
adv <- 42
```

```{r}

fstadv <- fstadv |>filter(StormKey == key, Adv <= adv)

```

## Probabilistic Storm Surge

Calculate the last date/time of forecast/advisory product and subtract 3 hrs

```{r}
dt <- (last(fstadv$Date) - (60 * 60 * 3)) |> 
    strftime(format = "%Y%m%d%H", tz = "UTC")
```

Wrap `gis_download` in `safely`. Not all products will exist for every storm or even every advisory. 

I'm downloading the `psurge` products for 5 feet since there are no other storm surge products available for Ike.

```{r}
dl <- purrr::safely(.f = gis_download)
gis_surge <- gis_prob_storm_surge(key, products = list("psurge" = c(5)), 
                                  datetime = dt) |> dl()

if (!is.null(gis_surge$error))
    message(gis_surge$error)
```

Generate a base plot of the Atlantic ocean.

```{r}
bp <- al_tracking_chart(color = "black", fill = "white", size = 0.1, res = 50)
```

Since we're dealing with a polygon shapefile, we can get the bounding box of the dataset.

```{r}
bbox <- bbox(gis_surge$result$al092008_2008091112_gt5)
```

Add a little cushion for the map inset.

```{r}
lat_min <- bbox[2,1] - 2
lat_max <- bbox[2,2] + 2
lon_min <- bbox[1,1] - 4
lon_max <- bbox[1,2] + 4
```

Build a map inset.

```{r}
bp_inset <- ggplotGrob(bp +
    geom_rect(mapping = aes(xmin = lon_min, xmax = lon_max,
                            ymin = lat_min, ymax = lat_max),
              color = "red", alpha = 0) +
    theme_bw() +
    theme(axis.title = element_blank(),
          axis.ticks = element_blank(),
          axis.text.x = element_blank(),
          axis.text.y = element_blank(),
          plot.margin = margin(0, 0, 0, 0, "pt")))
```

Modify original `bp` zoomed in on our area of interest.

```{r}
bp <- bp +
    coord_equal(xlim = c(lon_min, lon_max),
                ylim = c(lat_min, lat_max)) +
    scale_x_continuous(expand = c(0, 0)) +
    scale_y_continuous(expand = c(0, 0)) +
    labs(x = "Lon",
         y = "Lat",
         caption = sprintf("rrricanes %s", packageVersion("rrricanes")))
```

Combine `bp` and `bp_inset` to finalize initial base plot. `bp` will be a base plot without the inset. `bpi` will have the inset.

```{r}
bpi <- bp + annotation_custom(grob = bp_inset, xmin = lon_max - 5,
                             xmax = lon_max - 1, ymin = -Inf,
                             ymax = lat_min + 5)
```

Convert the SpatialPolygonsDataframe to a dataframe.

```{r}
shp_storm_surge <- shp_to_df(gis_surge$result$al092008_2008091112_gt5) 
```

Probability of storm surge greater than five feet.

```{r}
bpi + geom_point(data = shp_storm_surge, 
                aes(x = long, y = lat, color = ProbSurge05), size = 1) + 
    theme(legend.position = "bottom", 
          legend.box = "vertical") +
    labs(title = "Probabilistic Storm Surge > 5ft", 
         caption = sprintf("rrricanes %s", packageVersion("rrricanes")))
```
