---
author: "Tim Trice"
date: "`r strftime(Sys.time(), '%a %b %e, %Y %R')`"
output: 
    html_document:
        code_folding: hide
        toc: true
        toc_depth: 3
params:
    title: "`r as.character()`"
title: "`r params$title`"
---

```{r setup, include = FALSE}
opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, 
                      fig.width = 10, fig.height = 7, fig.align = "center", 
                      fig.path = sprintf("%s/%s/", storm_vars$year, 
                                         storm_vars$key))
```

## Current Storm Details

```{r current_position}
# If any of necessary datasets exist, copy base plot to current plot
if (any(exists("storm_pts"), exists("storm_lin"), exists("storm_ww")))
    cp <- bp

if (exists("storm_lin")) {
    p_storm_lin <- geom_line(data = storm_lin, aes(x = long, y = lat))
    cp <- cp + p_storm_lin
}

if (exists("storm_pts")) {

    p_storm_pts <- geom_point(data = storm_pts, 
                              aes(x = Lon, y = Lat, fill = Status, 
                                  shape = Status, size = Wind))
    cp <- cp + p_storm_pts
}

if (exists("storm_ww")) {
    p_storm_ww <- geom_line(data = storm_ww, 
                            aes(x = long, y = lat, color = TCWW), size = 2)
    cp <- cp + p_storm_ww
}

# Print current plot
if (exists("cp")) 
    cp + 
    scale_color_manual(values = "blue") + 
    scale_shape_manual(values = c(21, 21, 21, 21)) + 
    guides(shape = guide_legend(override.aes = list(size = 3))) + 
    guides(size = guide_legend(nrow = 1)) + 
    theme(legend.position = "bottom", 
          legend.box = "vertical") + 
    labs(title = "Track, Watches and Warnings",
         subtitle = sprintf("%s (Adv %s, %s)", storm_vars$name, storm_vars$adv,
                            strftime(storm_vars$date, "%b %e, %Y %R %Z")))
```

### Wind and Pressure Details

```{r wind_radius}
if (any(exists("windswath"), exists("storm_radii")))
    wr <- bp

if (exists("windswath")) {
    p_windswath <- geom_polygon(data = windswath, 
                                  aes(x = long, y = lat, 
                                      fill = as.factor(RADII)), 
                                  alpha = 0.75)
    wr <- wr + p_windswath
}

if (exists("storm_radii")) {
    p_storm_radii <- geom_polygon(data = storm_radii, 
                                  aes(x = long, y = lat, 
                                      fill = as.factor(RADII)), 
                                  alpha = 0.75)
    wr <- wr + p_storm_radii
}

# Print wind radius
if (exists("wr")) 
    wr + 
    geom_point(data = storm_pts %>% slice(n()), aes(x = Lon, y = Lat), 
               size = 1, color = "black")
```

#### Wind Profile

```{r wind_profile}
# Plot wind values
wind_press %>% ggplot(aes(x = Date, y = Wind)) + 
    geom_line() + 
    geom_point(aes(color = Status), size = 3) + 
    scale_y_continuous(name = "Wind (kts)") + 
    theme_bw() + 
    theme(legend.position = "bottom")
```

#### Pressure Profile

```{r pressure_profile}
# Plot pressure values
wind_press %>% ggplot(aes(x = Date, y = Pressure)) + 
    geom_line() + 
    geom_point(aes(color = Status), size = 3) + 
    scale_y_continuous(name = "Pressure (mb)") + 
    theme_bw() + 
    theme(legend.position = "bottom")
```

#### Wind/Pressure Relational Change

```{r wind_press_chng}
wind_press_cum %>% ggplot(aes(x = Date, y = Val, group = Var, color = Var)) + 
    geom_line(size = 2) + 
    theme_bw() + 
    theme(legend.position = "bottom", legend.title = element_blank()) + 
    labs(title = "Wind/Pressure Change", 
         subtitle = "",
         y = "")
```

### Public Advisory Statement

`r pub_adv`

## Forecast

```{r plot_forecast}
# If any of necessary datasets exist, copy base plot to current plot
if (any(exists("fcst_lin"), exists("fcst_pts"), exists("fcst_pgn")))
    fp <- bp

## ---- 5-day forecast line ----------------------------------------------------
if (exists("fcst_lin")) {
    p_fcst_lin <- geom_path(data = fcst_lin, aes(x = long, y = lat))
    fp <- fp + p_fcst_lin
}

## ---- 5-day forecast points --------------------------------------------------
if (exists("fcst_pts")) {
    p_fcst_pts <- geom_point(data = fcst_pts, 
                                 aes(x = LON.x, y = LAT.x, color = TCDVLP.x, 
                                     size = MAXWIND.x))
    fp <- fp + p_fcst_pts
}

## ---- 5-day forecast polygon -------------------------------------------------
if (exists("fcst_pgn")) {
    p_fcst_pgn <- geom_polygon(data = fcst_pgn, 
                                  aes(x = long, y = lat, 
                                      fill = as.factor(FCSTPRD)), 
                                  alpha = 0.3)
    fp <- fp + p_fcst_pgn
}

# Print forecast plot
if (exists("fp")) fp
```

```{r tbl_forecast, results = "markdown"}
kable(fcst, caption = sprintf("%s Forecast", storm_vars$name))
```

### Cumulative Wind Speed Probabilities

#### 0.5-Degree Resolution

```{r check_wsp, echo = FALSE}
# If any of necessary datasets exist, copy base plot to current plot
if (any(exists("wsp34_half"), exists("wsp50_half"), exists("wsp64_half")))
    wsp <- bp
```

```{r wsp34_half, fig.cap = "Wind Speed Probabilities (>= 34kts)"}
if (exists("wsp34_half")) {
    p_wsp34_half <- geom_point(data = wsp34_half, 
                               aes(x = coords.x1, y = coords.x2, 
                                   size = PWIND120.x, color = PWIND120.x))
    wsp + p_wsp34_half
}
```

```{r wsp50_half, fig.cap = "Wind Speed Probabilities (>= 50kts)"}
if (exists("wsp50_half")) {
    p_wsp50_half <- geom_point(data = wsp50_half, 
                               aes(x = coords.x1, y = coords.x2, 
                                   size = PWIND120.x, color = PWIND120.x))
    wsp + p_wsp50_half
}
```

```{r wsp64_half, fig.cap = "Wind Speed Probabilities (>= 64kts)"}
if (exists("wsp64_half")) {
    p_wsp64_half <- geom_point(data = wsp64_half, 
                               aes(x = coords.x1, y = coords.x2, 
                                   size = PWIND120.x, color = PWIND120.x))
    wsp + p_wsp64_half
}
```

Cumulative wind speed probabilities are the sum of wind speed probablities for every forecast period issued. Below is a table (if available) of wind speed probabilities for select locations throughout each forecast period (12 hours, 24 hours, 36, 48, 72, 96 and 120). 

```{r tbl_wndprb, results = "markdown"}
if (nrow(wndprb) > 0) {
    kable(wndprb, caption = "Wind Speed Probabilities")
} else {
    message("There are no wind speed probabilities with this advisory.")
}
```

## Past Advisories

```{r tbl_advisories, results = "markdown"}
kable(fstadv %>% select(Status, Adv, Date, Wind, Gust, Pressure, Eye), 
      caption = sprintf("%s Advisories", storm_vars$name))
```

## Storm Discussion

`r discus$Contents`

## About `rrricanes`

<span style="color:red">rrricanes is not intended for use in emergency situations.</span>

`rrricanes` is a R package designed to get current and archived real-time data on tropical storms and hurricanes. Contributors are strongly encouraged. View the [GitHub repository](https://github.com/timtrice/rrricanes) for more information.
