---
title: "rrricanes in Data World"
author: "Tim Trice"
date: "June 15, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Methods of loading storm data in `rrricanes` includes parsing the National Hurricane Center archives which can be time-consuming and unreliable (consistent timeout issues). The function `load_storm_data` returns post-scraped datasets from the `rrricanesdata` and is much faster. However, you have to load the entire dataset and this may not be convenient for some users.

A third option is [data.world](https://data.world/) which is automatically updated from the `rrricanesdata` repository. The benefit of data.world is that you can custom build queries to extract only the data you need. 

Using data.world requires the `dwapi` package which can be installed from CRAN.

```{r, eval = FALSE}
install.packages("dwapi")
```

You will also need to get [your API key](https://data.world/settings/advanced) in order to access the datasets. See the "quickstart" vignette for help using `dwapi`.

```{r, eval = FALSE}
vignette("quickstart", package = "dwapi")
```

This article will focus on building SQL queries to retrieve data from data.world. There are some limitations to the functionality. See [SQL on data.world](https://docs.data.world/tutorials/dwsql/) for more information.

```{r, message = FALSE}
library(dplyr)
library(dwapi)
library(rrricanes)
```

Configure `dwapi` to use your own API key. Then set the `ds` parameter to the `rrricanes` dataset at "timtrice/rrricanes".

```{r}
configure(auth_token = Sys.getenv("dwapi_visitor_key"))
ds <- "timtrice/rrricanes"
```

## storms

The `storms` dataframe is a summary table of all storms that exist in the dataset. 

```{r}
ds %>% sql(query = "SELECT * FROM storms LIMIT 5")
```

### Find a storm by name

```{r}
ds %>% sql(query = "SELECT * FROM storms WHERE Name = 'Katrina'")
```

### Aggregate storms by month

```{r}
ds %>% sql(query = "SELECT COUNT(Key) AS `n`,
        MONTH(CAST(storms.StartDate AS STRING)) AS `Mth`
    FROM `storms`
    GROUP BY MONTH(CAST(storms.StartDate AS STRING))
    ORDER BY `Mth`;")
```

## adv

The `adv` dataframe lists the core data of every advisory for every storm. This excludes forecast points and all wind radii.

```{r}
ds %>% sql(query = "SELECT * FROM adv LIMIT 5")
```

Earlier we saw there were two storms named Katrina. Let's build a query and return all the `adv` observations we have for Katrina (Key "AL122005").

```{r}
ds %>% sql(query = 
"SELECT * 
FROM adv
WHERE Key = 'AL122005';")
```

## wr

```{r}
ds %>% sql(query = "SELECT * FROM wr LIMIT 5")
```

```{r}
ds %>% sql(query = 
"SELECT * 
FROM wr
WHERE Key = 'AL122005';")
```

## fcst

```{r}
ds %>% sql(query = "SELECT * FROM fcst LIMIT 5")
```

```{r}
ds %>% sql(query = 
"SELECT * 
FROM fcst
WHERE Key = 'AL122005';")
```

## fcst_wr

```{r}
ds %>% sql(query = "SELECT * FROM fcst_wr LIMIT 5")
```

```{r}
ds %>% sql(query = 
"SELECT * 
FROM fcst_wr
WHERE Key = 'AL122005';")
```

## wndprb

```{r}
ds %>% sql(query = "SELECT * FROM wndprb LIMIT 5")
```

```{r}
ds %>% sql(query = 
"SELECT * 
FROM wndprb
WHERE Key = 'AL012006';")
```

## prblty

```{r}
ds %>% sql(query = "SELECT * FROM prblty LIMIT 5")
```

There was no `Key` value in strike probabilities. Therefore, must use a combination of `Status`, `Name` and, if necessary, `Adv` and `Date`.
